Effectiveness of community-based TURF-reserves in small-scale fisheries

### 2018-12-04 docs/Villasenor-Derbez.Rmd
Deleted unsupported part of sentence to close #12

We evaluate three community-based no-take marine reserves implemented in Mexican TURF-managed fisheries, therefore making them TURF-reserves (Fig \ref{fig:map}A). The first one was created by the *Buzos y Pescadores de la Baja California* fishing cooperative, located in Isla Natividad in the Baja California Peninsula (Fig \ref{fig:map}B). The main fishery in the island is the spiny lobster (*Panulirus interruptus*), but other resources like finfish, sea cucumber, read sea urchin, snail, and abalone are also an important source of income. In 2006, the community decided to implement two marine reserves within their fishing grounds to protect commercially important invertebrate species; mainly lobster and abalone. While these reserves obtained legal recognition only in 2018 \citep{dof_website_2018}, they have been well enforced since their implementation.

@@

These communities are representative of their region in terms of ecology, socioeconomic, and governance aspects. Isla Natividad, for example, is part of a greater group of fishing cooperatives belonging to a Federation of Fishing Cooperatives. This group has been identified as a cohesive group that often cooperates to better manage their resources \citep{mccay_2014-CN,mccay_2017,acevesbueno_2017}. Likewise, Maria Elena and Punta Herrero are representative of fishing cooperatives in the Mexican Caribbean, which are also part of a regional Federation. Together, these three communities provide an accurate representation of other fishing communities that have been historically manged with TURFs in each of their regions. While each region has additional communities that have established community-based TURF-reserves, available data would not allow us to perform the in-depth causal inference analysis that we undertake. Yet, given the similarities among communities and the socioeconomic and governance setting under which they operate, it is safe to cautiously generalize our insights to other similar reserves in Mexico and elsewhere around the world.

@@

We use a difference-in-differences analysis to evaluate these indicators. This approach allows us to estimate the effect that the reserve had by comparing trends across time and treatments \citep{moland_2013-VP,Villasenor-Derbez_2018}. The analysis of ecological indicators is performed with a multiple linear regression of the form:

>>>>

We evaluate three community-based no-take marine reserves implemented in Mexican TURF-managed fisheries, therefore making them TURF-reserves (Fig \ref{fig:map}A). The first one was created by the *Buzos y Pescadores de la Baja California* fishing cooperative, located in Isla Natividad in the Baja California Peninsula (Fig \ref{fig:map}B). The main fishery in the island is the spiny lobster (*Panulirus interruptus*), but other resources like finfish, sea cucumber, read sea urchin, snail, and abalone are also an important source of income. In 2006, the community decided to implement two marine reserves within their fishing grounds to protect commercially important invertebrate species; mainly lobster and abalone. These reserves obtained legal recognition only in 2018 \citep{dof_website_2018}.

@@

These communities are representative of their region in terms of ecology, socioeconomic, and governance aspects. Isla Natividad, for example, is part of a greater group of fishing cooperatives belonging to a Federation of Fishing Cooperatives. This group has been identified as a cohesive group that often cooperates to better manage their resources \citep{mccay_2014,mccay_2017,acevesbueno_2017}. Likewise, Maria Elena and Punta Herrero are representative of fishing cooperatives in the Mexican Caribbean, which are also part of a regional Federation. Together, these three communities provide an accurate representation of other fishing communities that have been historically manged with TURFs in each of their regions. While each region has additional communities that have established community-based TURF-reserves, available data would not allow us to perform the in-depth causal inference analysis that we undertake. Yet, given the similarities among communities and the socioeconomic and governance setting under which they operate, it is safe to cautiously generalize our insights to other similar reserves in Mexico and elsewhere around the world.

@@

We use a difference-in-differences analysis to evaluate these indicators. This approach allows us to estimate the effect that the reserve had by comparing trends across time and treatments \citep{moland_2013,Villasenor-Derbez_2018}. The analysis of ecological indicators is performed with a multiple linear regression of the form:

### 2018-12-04 docs/Villasenor-Derbez.tex
Deleted unsupported part of sentence to close #12

Coastal marine ecosystems provide livelihoods for small-scale fishers
and coastal communities around the world. Small-scale fisheries face
great challenges since they are difficult to monitor, enforce, and
manage. Combining territorial user rights for fisheries (TURF) with
no-take marine reserves to create TURF-reserves can improve the
performance of small-scale fisheries by buffering fisheries from
environmental variability and management errors, while ensuring that
fishers reap the benefits of conservation investments. In the last 12
years, 18 old and new community-based TURF-reserves gained legal
recognition thanks to a 2014 regulation; their effectiveness has not
been formally evaluated. We combine causal inference techniques and the
Social-Ecological Systems framework to provide a holistic evaluation of
community-based TURF-reserves in three coastal communities in Mexico. We
find that while reserves have not yet achieved their stated goal of
increasing the density of lobster and other benthic invertebrates, they
continue to receive significant support from the fishing communities. A
lack of clear ecological and socioeconomic effects likely results from a
combination of factors. First, some of these reserves might be too young
for the effects to show. Second, the reserves are not large enough to
protect mobile species, like lobster. Third, variable and extreme
oceanographic conditions have impacted harvested populations. Fourth,
local fisheries are already well managed, and it is unlikely that
reserves might have a detectable effect in landings. However, these
reserves have shaped small-scale fishers' way of thinking about marine
conservation, which can provide a foundation for establishing
additional, larger marine reserves needed to effectively conserve mobile
species.

@@

TURFs are a fisheries management tool in which a well defined group of
fishers have exclusive access to an explicitly delimited portion of the
ocean. They promote a sense of stewardship and incentivise resource
users to sustainably manage their resources
\citep{gelcich_2008,mccay_2017}. On the other hand, no-take marine
reserves (marine reserves from hereinafter) are areas where all
extractive activities are off-limits. These can be implemented to
protect biodiversity but also as fishery management tools that restrict
fishing effort and gears and therefore aid in the recovery of marine
stocks. Commonly known as TURF-reserves, the combination of these tools
should in theory increase the benefits of spatial access rights allowing
the maintenance of healthy resources
\citep{afflerbach_2014,lester_2017}.

Research on TURFs has shown that these areas have higher abundance of
targeted species than sites operating under open access
\citep{gelcich_2008,mccay_2014,mccay_2017}. Likewise, conservation
science has shown how marine reserves lead to increased biomass, species
richness, and abundance within the protected regions
\citep{lester_2009-Ks,giakoumi_2017-V2,sala_2017-69}, and that these may
have a series of additional benefits like climate change mitigation,
protection from environmental variability, and fisheries benefits
\cite{roberts_2017-J9,micheli_2012-EU,krueck_2017-J1}. The benefits
resulting from reserves established within TURFs should be captured
exclusively by the group of fishers with exclusive access. Although in
theory these systems are successful
\citep{costello_2010,smallhornwest_2018}, there is little empirical
evidence of their effectiveness and the drivers of their success
\citep{afflerbach_2014,lester_2017}.

TURF-reserve systems are inherently intrincate social-ecological
systems, and their effectiveness must depend on how environmental and
social factors combine and interact. This is especially important in
social-ecological coastal systems dominated by close interactions and
feedbacks between people and natural resources \citep{ostrom_2009-hg}.
There is a growing body of literature focusing on the relations between
socioeconomic and governance structures and reserve effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_,bergseth_2018}.
However, to our knowledge, no studies exist that evaluate TURF-reserves
from both a social and ecological perspective.

@@

Before discussing our data collection methods and describing our
analyses, our case studies warrant some more background. TURF-reserves
can be created as community-based marine reserves, voluntarily
established and enforced by local communities. This bottom-up approach
increases compliance and self-enforcement, and reserves can yield
benefits similar to systematically-designed reserves
\citep{gelcich_2015,beger_2004,smallhornwest_2018}. Community-based
spatial closures occur in different contexts, like the \emph{kapu} or
\emph{ra'ui} areas in the Pacific Islands
\citep{bohnsack_2004,johannes_2002}. However, marine reserves are
difficult to enforce if they are not legally recognized, and fishers
rely on the exclusive access granted by the TURF.

@@

\hypertarget{study-area}{%
\subsection{Study area}\label{study-area}}

We evaluate three TURF-reserves in Mexico (Fig \ref{fig:map}A). The
first one was created by the \emph{Buzos y Pescadores de la Baja
California} fishing cooperative, located in Isla Natividad in the Baja
California Peninsula (Fig \ref{fig:map}B). The main fishery in the
island is the spiny lobster (\emph{Panulirus interruptus}), but other
resources like finfish, sea cucumber, read sea urchin, snail, and
abalone are also an important source of income. In 2006, the community
decided to implement two marine reserves within their fishing grounds to
protect commercially important invertebrate species; mainly lobster and
abalone. While these reserves obtained legal recognition only in 2018
\citep{dof_website_2018}, they have been well enforced since their
implementation.

@@

These communities are representative of their region in terms of
ecology, socioeconomic, and governance aspects. Isla Natividad, for
example, is part of a greater group of fishing cooperatives belonging to
a Federation of Fishing Cooperatives. This group has been identified as
a cohesive group that often cooperates to better manage their resources
\citep{mccay_2014-CN,mccay_2017,acevesbueno_2017}. Likewise, Maria Elena
and Punta Herrero are representative of fishing cooperatives in the
Mexican Caribbean, which are also part of a regional Federation.
Together, these three communities provide an accurate representation of
other fishing communities in each of their regions. While each region
has additional communities that have established community-based
TURF-reserves, available data would not allow us to perform the in-depth
causal inference analysis that we undertake. Yet, given the similarities
among communities and the socioeconomic and governance setting under
which they operate, it is safe to cautiously generalize our insights to
other similar reserves in Mexico and elsewhere around the world.

@@

The model interpretation remains as for Eq \ref{eqn:reg_bio}, but in
this case the \(Treated\) dummy variable indicates if the community has
a reserve (\(Treated = 1\)) or not (\(Treated = 0\)) and \(\sigma_jCom\)
captures community-level fixed-effects. These regression models allow us
to establish a causal link between the implementation of marine reserves
and the observed trends by accounting for temporal and spatial dynamics
\citep{depalma_2018}. The effect of the reserve is captured by the
\(\lambda_t\) coefficient, and represents the difference observed
between the control site before the implementation of the reserve and
the treated sites at time \(t\) after controlling for other time and
space variations (\emph{i.e.} \(\gamma_t\) and \(\beta\) respectively).
All model coefficients were estimated via ordinary least-squares and
heteroskedastic-robust standard errors \citep{zeileis_2004-7n}. All
analyses were performed in R version 3.5.1 (2018-07-02) and R Studio
version 1.1.456 \citep{R_2018}.

@@

@@

Some works evaluate marine reserves by performing inside-outside
\citep{guidetti_2014-8Z,friedlander_2017-oI,rodriguez_2017-PD} or
before-after comparisons \citep{betti_2017-lq}. The first approach does
not address temporal variability, and the second can not distinguish
between the temporal trends in a reserve and the entire system
\citep{depalma_2018}. Our approach to evaluate the temporal and spatial
changes provides a more robust measure of reserve effectiveness. For
example, we capture previously described patterns like the rapid
increase observed for lobster densities in Isla Natividad on the sixth
year (\emph{i.e.} 2012; Fig. \ref{fig:indicators}A), a year after the
hypoxia events described by \citet{micheli_2012-EU}, which caused mass
mortality of sedentary organisms such as abalone and sea urchins, but
not lobster and finfish. Yet, our empirical approach assumes control
sites are a plausible counterfactual for treated sites. This implies
that treated sites would have followed the same trend as control sites,
had the reserves not been implemented. Nonetheless, temporal trends for
each site don't show any significant increases (S1 Table, S2 Table, S3
Table), supporting our findings of lack of change in the indicators
used.

>>>>

Coastal marine ecosystems provide livelihoods for small-scale fishers
and coastal communities around the world. Small-scale fisheries face
great challenges since they are difficult to monitor, enforce, and
manage. Combining territorial user rights for fisheries (TURF) with
no-take marine reserves to create TURF-reserves can improve the
performance of small-scale fisheries by buffering fisheries from
environmental variability and management errors, while ensuring that
fishers reap the benefits of conservation investments. In the last 12
years, 18 old and new community-based TURF-reserves gained legal
recognition thanks to a 2014 regulation; their effectiveness has not
been formally evaluated. We combine causal inference techniques and the
Social-Ecological Systems framework to provide a holistic evaluation of
community-based TURF-reserves in three coastal communities in Mexico. We
find that while reserves have not yet achieved their stated goal of
increasing the density of lobster and other benthic invertebrates, they
continue to receive significant support from the fishing communities. A
lack of clear ecological and socioeconomic effects likely results from a
combination of factors. First, some of these reserves might be too young
for the effects to show. Second, the reserves are not large enough to
protect mobile species, like lobster. Third, variable and extreme
oceanographic conditions have impacted harvested populations. Fourth,
local fisheries are already well managed, and it is unlikely that
reserves might have a detectable effect in landings. However, these
reserves may provide a foundation for establishing additional, larger
marine reserves needed to effectively conserve mobile species.

@@

TURFs are a fisheries management tool in which a well defined group of
fishers have exclusive access to an explicitly delimited portion of the
ocean. They promote a sense of stewardship and incentivise resource
users to sustainably manage their resources
\citep{gelcich_2008,mccay_2017}. On the other hand, no-take marine
reserves (marine reserves from hereinafter) are areas where all
extractive activities are off-limits. These can be implemented to
protect biodiversity but also as fishery management tools that restrict
fishing effort and gears and therefore aid in the recovery of marine
stocks. Commonly known as TURF-reserves, the combination of these tools
should in theory increase the benefits of spatial access rights allowing
the maintenance of healthy resources
\citep{afflerbach_2014,gelcich_2015,lester_2017}.

Research on TURFs has shown that these areas have higher abundance of
targeted species than sites operating under open access
\citep{gelcich_2008,mccay_2014,mccay_2017}. Likewise, conservation
science has shown how marine reserves lead to increased biomass, species
richness, and abundance within the protected regions
\citep{lester_2009,giakoumi_2017-V2,sala_2017-69}, and that these may
have a series of additional benefits like climate change mitigation,
protection from environmental variability, and fisheries benefits
\cite{roberts_2017-J9,micheli_2012-EU,krueck_2017-J1}. The benefits
resulting from reserves established within TURFs should be captured
exclusively by the group of fishers with exclusive access. Although in
theory these systems are successful
\citep{costello_2010,smallhornwest_2018}, there is little empirical
evidence of their effectiveness and the drivers of their success
\citep{afflerbach_2014,lester_2017}.

TURF-reserve systems are inherently intrincate social-ecological
systems, and their effectiveness must depend on how environmental and
social factors combine and interact \citep{gelcich_2015}. This is
especially important in social-ecological coastal systems dominated by
close interactions and feedbacks between people and natural resources
\citep{ostrom_2009}. There is a growing body of literature focusing on
the relations between socioeconomic and governance structures and
reserve effectiveness
\citep{halpern_2013,lpezangarita_2014,mascia_2017-m_,bergseth_2018}.
However, to our knowledge, no studies exist that evaluate TURF-reserves
from both a social and ecological perspective.

@@

\hypertarget{turf-reserves-in-mexico}{%
\subsection{TURF-reserves in Mexico}\label{turf-reserves-in-mexico}}

Before discussing our data collection methods and describing our
analyses, our case studies warrant some more background. Community-based
marine reserves that are implemented within TURFs are a form of
TURF-reserves, voluntarily established and enforced by local
communities. This bottom-up approach increases compliance and
self-enforcement, and reserves can yield benefits similar to
systematically-designed reserves
\citep{beger_2004,gelcich_2015,smallhornwest_2018}. Community-based
spatial closures occur in different contexts, like the \emph{kapu} or
\emph{ra'ui} areas in the Pacific Islands
\citep{bohnsack_2004,johannes_2002}. However, marine reserves are
difficult to enforce if they are not legally recognized, and fishers
rely on the exclusive access granted by the TURF.

@@

\hypertarget{study-areas}{%
\subsection{Study areas}\label{study-areas}}

We evaluate three community-based no-take marine reserves implemented in
Mexican TURF-managed fisheries, therefore making them TURF-reserves (Fig
\ref{fig:map}A). The first one was created by the \emph{Buzos y
Pescadores de la Baja California} fishing cooperative, located in Isla
Natividad in the Baja California Peninsula (Fig \ref{fig:map}B). The
main fishery in the island is the spiny lobster (\emph{Panulirus
interruptus}), but other resources like finfish, sea cucumber, read sea
urchin, snail, and abalone are also an important source of income. In
2006, the community decided to implement two marine reserves within
their fishing grounds to protect commercially important invertebrate
species; mainly lobster and abalone. These reserves obtained legal
recognition only in 2018 \citep{dof_website_2018}.

@@

These communities are representative of their region in terms of
ecology, socioeconomic, and governance aspects. Isla Natividad, for
example, is part of a greater group of fishing cooperatives belonging to
a Federation of Fishing Cooperatives. This group has been identified as
a cohesive group that often cooperates to better manage their resources
\citep{mccay_2014,mccay_2017,acevesbueno_2017}. Likewise, Maria Elena
and Punta Herrero are representative of fishing cooperatives in the
Mexican Caribbean, which are also part of a regional Federation.
Together, these three communities provide an accurate representation of
other fishing communities that have been historically manged with TURFs
in each of their regions. While each region has additional communities
that have established community-based TURF-reserves, available data
would not allow us to perform the in-depth causal inference analysis
that we undertake. Yet, given the similarities among communities and the
socioeconomic and governance setting under which they operate, it is
safe to cautiously generalize our insights to other similar reserves in
Mexico and elsewhere around the world.

The regulation governing the implementation of Fish refuges states that
these are fishery management tools intended to have biological or
socioeconomic benefits \citep{nom}. For this reason, the main portion of
our analyses focuses on a series of biological and socioeconomic
indicators that may respond to reserve implementation. However, the
effectiveness of conservation and fisheries management interventions
also depends on the social and governance structures in place. We
therefore incorporate a reduced version of the Social Ecological Systems
framework \citep{ostrom_2009} and evaluate variables and indicators
known to aid and hinder the effectiveness of management interventions in
conservation and fisheries. The incorporation of the SES is not intended
to relate different levels of governance with reserve effectiveness, but
rather provide a discussion of potential causes of individual reserve
performance. The following two sections describe our data collection
methods and analyses of biological and socioeconomic indicators as well
as the SES analysis.

@@

The model interpretation remains as for Eq \ref{eqn:reg_bio}, but in
this case the \(Treated\) dummy variable indicates if the community has
a reserve (\(Treated = 1\)) or not (\(Treated = 0\)) and \(\sigma_jCom\)
captures community-level fixed-effects. These regression models allow us
to establish a causal link between the implementation of marine reserves
and the observed trends by accounting for temporal and spatial dynamics
\citep{depalma_2018}. The effect of the reserve is captured by the
\(\lambda_t\) coefficient, and represents the difference observed
between the control site before the implementation of the reserve and
the treated sites at time \(t\) after controlling for other time and
space variations (\emph{i.e.} \(\gamma_t\) and \(\beta\) respectively).
All model coefficients were estimated via ordinary least-squares and
heteroskedastic-robust standard errors \citep{zeileis_2004-7n}. All
analyses were performed in R version 3.5.1 and R Studio version 1.1.456
\citep{R_2018}. All data and code ara available in a
\href{https://github.com/jcvdav/ReserveEffect}{GitHub repository}.

@@

The fact that these reserves were re-established and just enacted shows
a comitment of support (issue \#16 on GH). the SES provides a static
picture of governance and resource conditions, whereas the biological
assessment reflects changes through time (Isse \#31).

@@

Some works evaluate marine reserves by performing inside-outside
\citep{guidetti_2014-8Z,friedlander_2017-oI,rodriguez_2017-PD} or
before-after comparisons \citep{betti_2017-lq}. The first approach does
not address temporal variability, and the second can not distinguish
between the temporal trends in a reserve and the entire system
\citep{depalma_2018}. While many ecology studies have used BACI sampling
designs and respective analyses (\emph{e.g.} \citep{stewartoaten_1986}),
few conservations studies have done so to evaluate the effect of an
intervention (\emph{i.e.}
\citet{francinifilho_2008,lester_2009,moland_2013}) which has resulted
in a call for more robust analyses in conservation science
\citep{guidetti_2002,ferraro_2006}.

Our approach to evaluate the temporal and spatial changes provides a
more robust measure of reserve effectiveness. For example, we capture
previously described patterns like the rapid increase observed for
lobster densities in Isla Natividad on the sixth year (\emph{i.e.} 2012;
Fig. \ref{fig:indicators}A), a year after the hypoxia events described
by \citet{micheli_2012-EU}, which caused mass mortality of sedentary
organisms such as abalone and sea urchins, but not lobster and finfish.
Yet, our empirical approach assumes control sites are a plausible
counterfactual for treated sites. This implies that treated sites would
have followed the same trend as control sites, had the reserves not been
implemented. Nonetheless, temporal trends for each site don't show any
significant increases (S1 Table, S2 Table, S3 Table), supporting our
findings of lack of change in the indicators used.
